---
title: "p8105_hw2_sef2183"
author: "Sarah Forrest"
date: "2022-09-27"
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

```{r load packages, echo = FALSE}
library(tidyverse)
```

# Problem 1
## Read and Clean the Data
```{r problem 1 - read and manipulate data}
subway_df = 
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada) %>%
  mutate(entry = recode(entry, "YES" = TRUE,  "NO" = FALSE))

subway_df

is.logical(subway_df$entry)
```

## Dataset Description: 
This dataset contains the variables: line, station_name, station_latitude, station_longitude, routes served (route1-route11), entry, vending, entrance_type, and ada (ADA compliance). 

So far, I imported the CSV file with the data using the `read_csv` function and a relative path to the data file. Then, I used the pipe operator to feed the data frame into the `janitor` package's `clean_names` function to  convert all variable names to lower case and put underscores in the gaps Then, I used the pipe operator again to feed the output of the `clean_names` function to the `select` function to retain only the variables of interest in the data frame. Then, I used the pipe operator again to feed the output of the `select` function into the `recode` function within `mutate` to convert the entry variable from character (`YES` vs `NO`) to a logical variable. Finally, the outcome of all of these processes are stored in the tibble `subway_df`. I printed out the data set to check the variables that were included in it and also checked that the entry variable was logical using the `is.logical` function which returned "TRUE".

The dimensions of the resulting `subway_df` tibble is 1,868 rows (observations) by 19 columns (variables).

These data are not tidy. Variables route1-route11 span 11 columns in the dataset. In order to be tidy, there should just be 1 column for the variable "route_number" with values for each route number and another column, "route_name" for the name of the corresponding subway.

## Dataset Questions:
```{r}
distinct_stations_df = distinct(subway_df, station_name, line, .keep_all = TRUE)

nrow(distinct_stations_df)
```

* There are 465 distinct stations.

```{r}
sum(distinct_stations_df$ada == "TRUE")
```

* 84 distinct stations are ADA compliant. 

```{r}
denom_df = filter(distinct_stations_df, vending == "NO")

denom = sum(distinct_stations_df$vending == "NO")
num = sum(denom_df$entry == "TRUE")

num / denom
```

* 5/9 station entrances / exits without vending allow entrance

```{r problem 1 - reformat data}
summary(subway_df)

subway_df$route8 <- as.character(subway_df$route8)
subway_df$route9 <- as.character(subway_df$route9)
subway_df$route10 <- as.character(subway_df$route10)
subway_df$route11 <- as.character(subway_df$route11)

summary(subway_df)

subway_long_df =
  subway_df %>%
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name")
  
subway_long_df
```

```{r}
distinct_stations_long_df = distinct(subway_long_df, station_name, line, .keep_all = TRUE)

sum(distinct_stations_long_df$route_name == "A")
```

60 distinct stations serve the A train. 

```{r}
a_df = filter(distinct_stations_long_df, route_name == "A") 
sum(a_df$ada == "TRUE")
```

17 of the stations that serve the A train are ADA compliant.

# Problem 2
```{r problem 2 - read and manipulate data}
Mr_Trash_Wheel_df =
  read_excel = ("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx") %>%
  janitor::clean_names() %>%
```

specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
use reasonable variable names
omit rows that do not include dumpster-specific data
round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
Use a similar process to import, clean, and organize the data for Professor Trash Wheel, and combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

## Dataset Description: 
You are encouraged to use inline R. 
The number of observations in the resulting dataset is ___.
Examples of key variables in the dataset include: _____. 
For available data, the total weight of trash collected by Professor Trash Wheel is ____.
The total number of sports balls collected by Mr. Trash Wheel in 2020 was ___.

# Problem 3

```{r problem 3 - read and manipulate data}
pols_month_df = 
  read_csv("data/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate()

snp_df = 
  read_csv("data/fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names() %>%
  separate()

unemployment_df = 
  read_csv("data/fivethirtyeight_datasets/unemployment.csv")
```

Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

Join the datasets by merging snp into pols, and merging unemployment into the result.

## Dataset Descriptions: 
The pols-month dataset contained:

The snp dataset contained:

The unemployment dataset contained: 

The resulting dataset contains (dimension, range of years, names of key variables):


